### 2023-03-26 definitions.tex
details

\begin{fact}\label{stop condition}
% Fact 4 (stop condition). 
If $\mathscr{S}$ is a WSTS with strong monotony and $U \subseteq S$ is an upward-closed set and $k \geq 0$ s.t. $U^k = U^{k+1}$ , then $U^\ell = U^k$ for all $\ell \geq k$, i.e.,
$k(U) \leq k$. This also implies that $\pred^*(U) = U^k$.
\end{fact}

@@

Since $\pred(\uparrow \{s' \}$) is computable for any $s'\in S$ by definition, we obtain a finite generating set of $U^{k+1}$ . By
Fact~\ref{fact basis}, we can compute a basis of $U^{k+1}$.
\end{proof}

>>>>

\begin{fact}\label{stop condition}
% Fact 4 (stop condition). 
If $\mathscr{S}$ is a WSTS with strong compatibility and $U \subseteq S$ is an upward-closed set and $k \geq 0$ s.t. $U^k = U^{k+1}$ , then $U^\ell = U^k$ for all $\ell \geq k$, i.e.,
$k(U) \leq k$. This also implies that $\pred^*(U) = U^k$.
\end{fact}

@@

Since a finite basis of $\pred(\uparrow \{s' \}$) is computable for any $s'\in S$ by definition, we obtain a finite generating set of $U^{k+1}$ . By
Fact~\ref{fact basis}, we can compute a basis of $U^{k+1}$.
\end{proof}

### 2023-03-26 resilience.tex
details

The resilience problem (resp. the $k$-resilience problem) in a transition system is to decide whether from a bad state, there exists a path (resp. a path of length smaller than $k$) that reaches a safe state. $\Safe, \Bad \subseteq S$ and we suppose that $\Safe \cap  \Bad = \emptyset$. A usual case is $Safe = S \backslash Bad$ and then Safe is downward-closed iff Bad is upward-closed. Let us formalize these properties.

\problemx{resilience problem (RP)}
{A transition system $\mathscr{S}=(S,\rightarrow)$ and two sets $\Safe, \Bad \subseteq S$.}
{$\Bad \longrightarrow^{*} \Safe$ ?\newline}
%
%\alain{il faudrait ne pas répéter 3 fois les mêmes imputs pour les 3 pbs: énoncer les 3 uniformes pbs d'un coup avec une fois l'input puis les 3 pbs pour un état s donné sans répéter non plus 3 fois les mêmes inputs}

@@

\problemx{$k$-resilience problem (kRP)}
{A transition system $\mathscr{S}=(S,\rightarrow), k \in \mathbb{N}$ and two sets $\Safe, \Bad \subseteq S$.}
{$\Bad \longrightarrow^{\leq k} \Safe$ ?\newline}

\problemx{bounded resilience problem (BRP)}
{A transition system $\mathscr{S}=(S,\rightarrow)$ and two sets $\Safe, \Bad \subseteq S$.}
%{$\exists k \geq 0 ~ \forall s' \in D ~ s \rightarrow^* s' \implies \exists s'' \in U ~ s' \rightarrow^{\leq k} s''$ ?\newline}
{$\exists k \geq 0$ such that $\mathscr{S}$ is %uniformely
 $k$-resilient ?\newline}

\alain{trouver les \Bad~ et \Safe~ maximum tels que S est resilient. est-ce vrai que si S est $(B_i,D_i)$-resilient alors S est $(\cap, \cup B_i,D_i)$-resilient ? on pourrait décider que \Safe=complement de \Bad~ et donc on aurait un seul ensemble (et son complement)}

These three reachability problems are decidable for finite transition systems but undecidable for (general) infinite-state transition systems. 
So we restrict our framework to the class of infinite-state WSTS. Since most of decidable properties in WSTS are given as effective upward or downward closed sets \cite{DBLP:journals/iandc/AbdullaCJT00, DBLP:journals/tcs/FinkelS01}, we consider upward closed or downward closed sets $\Safe$ and $\Bad$.
\alain{on peut penser à des ensembles $\Bad$ definis dans une logique booleennne sur les clos par le bas, haut, +...}

For instance, the well-known mutual exclusion property is often modelized in a $d$-counters machine by the property that a counter $c_{mutex}$ must be bounded by (usually) one. Then, the set $\Safe =  \{c_{mutex} \leq 1\} \times \mathbb{N}^{d-1}$ is downward closed and $\Bad =\{c_{mutex} \geq 2\} \times  \mathbb{N}^{d-1} $ is the upward closed complementary of $\Safe$. In \cite{DBLP:conf/gg/Ozkan22}, the authors considered that $\Bad$ is always downward closed and $\Safe$ is always upward closed.
%		
RP is decidable for lossy counter machines (LCM) with $\Safe$ and $\Bad$ semilinear sets as a consequence of results in Section 3.4 of \cite{DBLP:conf/rp/Schnoebelen10}. In particular, the existing proofs of resilience use the decidability of reachability; but we wish to decide resilience for models with undecidable reachability.
%
In what follows, we will always suppose that a downward closed set $D$ is given by its finite set of ideals $I_D=\{J_1, J_2,...,J_n\}$ satisfying $D=\downarrow D = \downarrow (J_1 \cup J_2 \cup..\cup J_n)$ and that an upward closed set $U$ is given by its finite set $Min_U$ of its minimal elements satisfying $U=\uparrow Min_U$.

@@

\begin{proof}
Since $\Safe=\uparrow \Safe$ and
$\mathscr{S}=(S,\rightarrow,\leq)$ is a WTSTS with strong %upward-
compatibility, then $\pred^k(\Safe)= \uparrow~\pred^k(\Safe)$ for all $k \in \N$, and $\pred^k(\Safe)$ admits a finite basis $B_{\pred^k(\Safe)}$. From a similar reasoning as above, the $k$-resilience problem is decidable.\alain{réserver k pour le k-résilience, il y a d'autres lettres...n, m, p,...}
\end{proof}

@@

\begin{proof}
Let $\{J_1, J_2,...,J_n\}$ be the ideal decomposition of $\Bad$ and $B_{\Safe}=\{b_1,b_2,...,b_m\}$ be the minimal basis of $\Safe$.
The %uniform 
resilience problem (RP) is equivalent to an infinite number of instances of the coverability problem : for all $x \in \Bad$ does there exist an $j$ such that $b_j$ is coverable from $x$. This infinite set of coverability questions can be reduced to a \emph{finite} number of instances of the coverability problem in the completion $\hat{\mathscr{S}}=(Ideals(S),\rightarrow, \subseteq)$ of $\mathscr{S}=(S,\rightarrow, \leq)$.

@@

Let us show that $b_j$ is coverable from $x$ in $\mathscr{S}$ if and only if $\downarrow b_j$ is coverable (for inclusion) from $\downarrow x$ in $\hat{\mathscr{S}}$.

Suppose that $b_j$ is coverable from $x$ then there exists a run $x \xrightarrow{k} y \geq b_j$. From Proposition 30, there exist an ideal $J$ and a run $\downarrow x \xrightarrow{k} J$ where $J \supseteq \downarrow y \supseteq \downarrow b_j$ in $\hat{\mathscr{S}}$, hence $\downarrow b_j$ is covered from $\downarrow x$.
Conversely, if $I \xrightarrow{k} J$ in $\hat{\mathscr{S}}$ with $\downarrow b_j \subseteq J$ then 
%	for every $y \in J$, 
there exists $x \in I$ and $y' \geq b_j$ such that $x \xrightarrow{k} y'  \geq b_j$ in $\mathscr{S}$ and then $b_j$ is coverable from $x$ in $\mathscr{S}$.
%

This reduction to a finite number of questions implies that resilience and bounded resilience are equivalent since, here, resilience implies bounded resilience (bounded resilience always implies resilience): if it is resilient then it is $k$-resilient for $k$ the maximum of lengths of covering sequences.
%
% iff there is a run in $\hat{S}$ from an ideal I to J such that  $\downarrow x \subseteq I$ and $\downarrow b_j \subseteq J$. This is a consequence of  
%Propositions 29 and 30 in \cite{BFM-icalp14} that establish a strong relation between the runs of a WSTS $\mathscr{S}=(S,\rightarrow, \leq)$ with its completion $\hat{S}$.
%' \in I(\Bad)$.

To decide the bounded resilience, we decide $n \times m$ coverability questions: is state $b_j$ coverable from ideal $J_i$ ? If all these $n \times m$ coverability questions are positive then we compute $K=max(k_{i,j} \mid i=1,..,n $ and $j= 1,..m)$ where $k_{i,j}$ is the least length of a sequence that covers  $b_j$ from $J_i$.
\alain{peut-être definir $K_{ \mathscr{S}}(Safe,Bad)$ en dehors de la preuve puisqu'on l'utilise ailleurs (théorème suivant) comme suit: If all these $n \times m$ coverability questions are positive then  $K(Safe,Bad)=max(k_{i,j})$ else $K=\infty$}
 % \alain{comment trouver les $k_a$ ?} 
 by iteratively checking the different possibles values of every $k_{i,j}$ until we find the minimal one,
 and we deduce that  $\mathscr{S}$ is $K$-resilient,
%  \alain{else $\mathscr{S}$ is not $K$-resilient...et alors qu'en déduit-on ? il pourrait exister un $K' \geq K$ pour lequel $\mathscr{S}$ est K'-resilient....}. 
else if some of these $n \times m$ coverability questions are negative then resilience
do not hold and bounded resilience do not hold either.
 Since coverability is decidable for completion-post-effective $\omega^2$-WSTS, we have shown that both the 
  resilience problem (RP) and the  
  bounded resilience problem (BRP) are decidable.

@@

	\begin{proof}
\fi

We begin to compute $K$ and $n$ such that $ \uparrow \pred^n(\Safe)=  \uparrow \pred^*(\Safe)$.
If $K=\infty$ then $\mathscr{S}$ is not resilient for Safe and Bad.
Now, if $k \geq min(K,n)$, we conclude that is $k$-resilient; 
%	if moreover $\mathscr{S}=(S,\rightarrow, \leq)$ has the predbasis hypothesis, 
%	and then kRP is decidable.
	If $k <  min(K,n)$, then we check every path of length smaller
	than $min(K,n)$ from the ideals of the decomposition of $\Bad$ in the completion \alain{à expliquer mieux}.
\end{proof}

@@

>>>>

In a transition system $\mathscr{S}=(S,\rightarrow)$, we consider two subsets of states $\Safe, \Bad \subseteq S$ such that $\Safe \cap  \Bad = \emptyset$.
%	The property of mutual exclusion is often modelised with $\Safe = \downarrow \Safe$.
The \emph{resilience problem} (resp. the \emph{$k$-resilience problem}) is to decide whether from \emph{any} state in $\Bad$, \emph{there exists} a path (resp. a path of length smaller than or equal to $k$) that reaches a state in $\Safe$. We use the notation $\Bad \longrightarrow^{*} \Safe$ (resp. $\Bad \longrightarrow^{\leq k} \Safe$) for $\forall x \in \Bad, \exists y \in \Safe$ such that $x \longrightarrow^{*} y$ (resp.  $\forall x \in \Bad, \exists y \in \Safe$ such that $x \longrightarrow^{\leq k} y$). In our framework, $\Safe, \Bad \subseteq S$  are possibly infinite but they must admit a computable finite representation : for example, downward-closed sets and upward-closed sets in wqo and semilinear sets in $\mathbb{N}^d$ have finite basis. 
%	If $\Safe = S \backslash \Bad$, then $\Safe$ is downward-closed iff $\Bad$ is upward-closed. 
Let us formalize three resilience problems.

\problemx{resilience problem (RP)}
{A transition system $\mathscr{S}=(S,\rightarrow)$ and two finite encodings of sets $\Safe, \Bad \subseteq S$.}
{$\Bad \longrightarrow^{*} \Safe$ ?\newline}
%
%\alain{il faudrait ne pas répéter 3 fois les mêmes imputs pour les 3 pbs: énoncer les 3 uniformes pbs d'un coup avec une fois l'input puis les 3 pbs pour un état s donné sans répéter non plus 3 fois les mêmes inputs}

@@

\problemx{$k$-resilience problem (kRP)}
{A transition system $\mathscr{S}=(S,\rightarrow), k \in \mathbb{N}$ and two finite encodings of sets $\Safe, \Bad \subseteq S$.}
{$\Bad \longrightarrow^{\leq k} \Safe$ ?\newline}

\problemx{bounded resilience problem (BRP)}
{A transition system $\mathscr{S}=(S,\rightarrow)$ and two finite encodings of sets $\Safe, \Bad \subseteq S$.}
%{$\exists k \geq 0 ~ \forall s' \in D ~ s \rightarrow^* s' \implies \exists s'' \in U ~ s' \rightarrow^{\leq k} s''$ ?\newline}
{$\exists k \geq 0$ such that $\mathscr{S}$ is %uniformely
 $k$-resilient ?\newline}

These three reachability problems are decidable for finite transition systems but undecidable for (general) infinite-state transition systems. 
So we restrict our framework to the class of infinite-state WSTS. Since most of decidable properties in WSTS are given as effective upward or downward closed sets \cite{DBLP:journals/iandc/AbdullaCJT00, DBLP:journals/tcs/FinkelS01}, we consider upward closed or downward closed sets $\Safe$ and $\Bad$.

For instance, the well-known mutual exclusion property is often modelized in a $d$-counters machine by the property that a counter $c_{mutex}$ must be bounded by (usually) one. Then, the set $\Safe =  \{c_{mutex} \leq 1\} \times \mathbb{N}^{d-1}$ is downward closed and $\Bad =\{c_{mutex} \geq 2\} \times  \mathbb{N}^{d-1} $ is the upward closed complementary of $\Safe$.

In \cite{DBLP:conf/gg/Ozkan22}, the authors considered that $\Bad$ is downward closed and $\Safe$ is upward closed.
%		
RP is decidable for lossy counter machines (LCM) with $\Safe$ and $\Bad$ semilinear sets as a consequence of results in Section 3.4 of \cite{DBLP:conf/rp/Schnoebelen10}. In particular, the existing proofs of resilience use the decidability of reachability; but we wish to decide resilience for models with undecidable reachability.
%
In what follows, we will always suppose that a downward closed set $D=\downarrow D$ is given by its finite set of ideals $I_D=\{J_1, J_2,...,J_n\}$ satisfying $D = \downarrow (J_1 \cup J_2 \cup..\cup J_n)$ and that an upward closed set $U$ is given by its finite set $Min_U$ of its minimal elements satisfying $U=\uparrow Min_U$.

@@

\begin{proof}
Since $\Safe=\uparrow \Safe$ and
$\mathscr{S}=(S,\rightarrow,\leq)$ is a WTSTS with strong %upward-
compatibility, then $\pred^n(\Safe)= \uparrow~\pred^n(\Safe)$ for all $n \in \N$, and $\pred^n(\Safe)$ admits a finite basis $B_{\pred^n(\Safe)}$. From a similar reasoning as above, the $k$-resilience problem is decidable.
\alain{écrire une preuve}
%	\alain{réserver k pour le k-résilience, il y a d'autres lettres...n, m, p,...}
\end{proof}

@@

\begin{proof}
Let $\{J_1, J_2,...,J_n\}$ be the ideal decomposition of $\Bad$ and $\{b_1,b_2,...,b_m\}$ be the (unique) minimal basis of $\Safe$.
The %uniform 
resilience problem (RP) can be reduced to the following infinite number of instances of the coverability problem in $\mathscr{S}$: for all $x \in \Bad$ does there exist an $j$ such that $b_j$ is coverable from $x$. Let us show how this infinite set of coverability questions can be reduced to a \emph{finite} set of coverability questions in the completion $\hat{\mathscr{S}}=(Ideals(S),\rightarrow, \subseteq)$ of $\mathscr{S}=(S,\rightarrow, \leq)$.

@@

Let us prove that $b_j$ is coverable from $x$ in $\mathscr{S}$ if and only if $\downarrow b_j$ is coverable (for inclusion) from $\downarrow x$ in $\hat{\mathscr{S}}$.
%
Suppose that $b_j$ is coverable from $x$ then there exists a run $x \xrightarrow{k} y \geq b_j$. From Proposition 30, there exist an ideal $J$ and a run $\downarrow x \xrightarrow{k} J$ where $J \supseteq \downarrow y \supseteq \downarrow b_j$ in $\hat{\mathscr{S}}$, hence $\downarrow b_j$ is covered from $\downarrow x$.
Conversely, if $I \xrightarrow{k} J$ in $\hat{\mathscr{S}}$ with $\downarrow b_j \subseteq J$ then 
%	for every $y \in J$, 
there exists $x \in I$ and $y' \geq b_j$ such that $x \xrightarrow{k} y'  \geq b_j$ in $\mathscr{S}$ and then $b_j$ is coverable from $x$ in $\mathscr{S}$.

Hence we obtain: $\mathscr{S}$ is resilient iff for all $i=1,..,n$ and $j= 1,..m$, $\downarrow b_j$ is coverable from ideal $J_i$ in $\hat{\mathscr{S}}$.
%
Let us denote by $k_{i,j}$ the length of a covering sequence that covers $\downarrow b_j$ from $J_i$ in $\hat{\mathscr{S}}$ and let $k_{i,j}\stackrel{\text{def}}{=}\infty$ if $\downarrow b_j$ is not coverable from $J_i$. Let us now define $K_{\mathscr{S}}(\Safe,\Bad)=max(k_{i,j} \mid i=1,..,n$ and $j= 1,..m$).
%	(if all $k_{i,j}$ are finite) else $K=\infty$.
We now have $\mathscr{S}$ is resilient iff $K_{\mathscr{S}}(\Safe,\Bad)$ is finite iff $\mathscr{S}$ is $K_{\mathscr{S}}(\Safe,\Bad)$-resilient with $K_{\mathscr{S}}(\Safe,\Bad)$ finite.

This implies that resilience and bounded resilience are equivalent to coverability.
%
% iff there is a run in $\hat{S}$ from an ideal I to J such that  $\downarrow x \subseteq I$ and $\downarrow b_j \subseteq J$. This is a consequence of  
%Propositions 29 and 30 in \cite{BFM-icalp14} that establish a strong relation between the runs of a WSTS $\mathscr{S}=(S,\rightarrow, \leq)$ with its completion $\hat{S}$.
%' \in I(\Bad)$.
%
%To decide the bounded resilience, we decide $n \times m$ coverability questions: is state $\downarrow b_j$ coverable from ideal $J_i$ ? If all these $n \times m$ coverability questions are positive then we compute $K=max(k_{i,j} \mid i=1,..,n $ and $j= 1,..m)$ where $k_{i,j}$ is the least length of a sequence that covers  $\downarrow b_j$ from $J_i$.
%
 % \alain{comment trouver les $k_a$ ?} 
%
%  \alain{else $\mathscr{S}$ is not $K$-resilient...et alors qu'en déduit-on ? il pourrait exister un $K' \geq K$ pour lequel $\mathscr{S}$ est K'-resilient....}. 
%else if some of these $n \times m$ coverability questions are negative then resilience
%do not hold and bounded resilience do not hold either.
Since coverability is decidable for completion-post-effective $\omega^2$-WSTS, we deduce that both the 
  resilience problem (RP) and the bounded resilience problem (BRP) are decidable.

@@

	\begin{proof}

We begin to compute $K$ and $n$ such that $ \uparrow \pred^n(\Safe)=  \uparrow \pred^*(\Safe)$.
If $K=\infty$ then $\mathscr{S}$ is not resilient for Safe and Bad.
Now, if $k \geq min(K,n)$, we conclude that is $k$-resilient; 
%	if moreover $\mathscr{S}=(S,\rightarrow, \leq)$ has the predbasis hypothesis, 
%	and then kRP is decidable.
\fi

Let us now show that the $k$-resilience problem (kRP), with $k \in \mathbb{N}$, is also decidable.
Let us denote by $k'_{i,j}$ the \emph{minimal} length of a covering sequence that covers $\downarrow b_j$ from $J_i$ in $\hat{\mathscr{S}}$ if it exists and let $k'_{i,j}\stackrel{\text{def}}{=}\infty$ if $\downarrow b_j$ is not coverable from $J_i$. 
If $\downarrow b_j$ is coverable from $J_i$, we first compute an $k_{i,j}$, and then we compute $k'_{i,j}$ by iteratively checking whether there exists a sequence of length $0,1,...,k_{i,j}-1$ that covers $\downarrow b_j$ from $J_i$ until we find the minimal one which is necessarly smaller (or equal to) than $k_{i,j}$.

Let us now define $K'_{\mathscr{S}}(\Safe,\Bad)=max(k'_{i,j} \mid i=1,..,n$ and $j= 1,..m$) and we deduce that  $\mathscr{S}$ is $k$-resilient iff $k \geq K'_{\mathscr{S}}(\Safe,\Bad)$.

%	If $k <  min(K,n)$, then we check every path of length smaller
%	than $min(K,n)$ from the ideals of the decomposition of $\Bad$ in the completion \alain{à expliquer mieux}.
\end{proof}

@@

\alain{trouver les \Bad~ et \Safe~ maximum tels que S est resilient. est-ce vrai que si S est $(B_i,D_i)$-resilient alors S est $(\cap, \cup B_i,D_i)$-resilient ?}
%
\alain{on peut penser à des ensembles $\Bad$ definis dans une logique booleennne sur les clos par le bas, haut, +...}
